A Multi-Scale Analysis of 27,000 Urban Street Networks\footnote{This is a journal article manuscript currently under peer review.}

### 2017-03-28 latex/paper.tex
update citations

\usepackage{graphicx}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{pdflscape}
\usepackage[numbers]{natbib}
\usepackage[affil-it]{authblk}

\bibliographystyle{apalike}

@@

OpenStreetMap is an emerging source of worldwide geospatial data useful to researchers. This paper uses the OSMnx software to automatically download and analyze 27,000 street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---that is, every urbanized area, incorporated place, and Zillow neighborhood in the US. In turn it presents wide-ranging empirical findings on US urban form and street network characteristics. This paper addresses past limitations by conducting an analysis of street networks with an extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using non-planar, directed graphs. Its analysis emphasizes urban street network measures relevant to urban design and morphology, including structure, connectedness, centrality, and robustness. In total we cross-sectionally analyze the street networks of 497 urbanized areas, 19,655 cities and towns, and 6,857 neighborhoods. These sample sizes are larger than those in any previous study and illustrate the use of OSMnx and OpenStreetMap as a new urban form research platform.

@@

On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States Midwest and Great Plains available for free to all applicants \cite{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 270 million acres of public land (10\% of the entire US landmass) to private owners in the form of 1.6 million homesteads \cite{lee_kansas_1979, sherraden_inclusion_2005}. Towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \cite{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks.

A cross-sectional analysis of American urban form can reveal these artifacts and histories through street networks at metropolitan, municipal, and neighborhood scales. Network analysis is a natural approach to the study of cities as complex systems \cite{masucci_random_2009}. The empirical literature on street network analysis is growing ever richer, but suffers from some limitations---discussed in detail in \cite{boeing_osmnx:_2017} and summarized here. First, sample sizes have tended to be fairly small due to data availability, gathering, and processing constraints. Second, reproducibility has been difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability, or use dual graphs despite the loss of geographic and metric information.

This paper addresses these limitations by conducting an urban morphological analysis of street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using non-planar, directed graphs. In particular, it examines urban street network---represented here as primal, non-planar, weighted multidigraphs with possible self-loops---focusing on structure, connectedness, centrality, and resilience. Most studies in the street network literature that conduct topological and/or metric analysis tend to have sample sizes ranging around 10 to 50 networks. This study instead analyzes over 27,000 urban street networks at multiple overlapping scales across the United States. Namely, it examines the street networks of every incorporated city and town (n=19,655), census urbanized area (n=497), and Zillow-defined neighborhood (n=6,857) in the US. To do so, it uses OSMnx \cite{boeing_osmnx:_2017}---a new street network research toolkit---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales.

The purpose of this paper is threefold. First, it demonstrates a new methodology for easily and consistently acquiring, constructing, and analyzing large samples of street networks as nonplanar directed graphs. Second, it presents empirical findings of descriptive urban morphology for the street networks of every city, urbanized area, and Zillow neighborhood in the US. Third, it investigates with large sample sizes some claims previously made in the research literature with small sample sizes. This paper is organized as follows. In the next section, it discusses the data sources, tools, and methods used to collect, construct, and analyze these street networks. Next, it presents findings of the analyses at metropolitan, municipal, and neighborhood spatial scales. Finally, it concludes with a discussion of these findings and their implications for street network analysis, urban morphology, and city planning.

\section{Methodology}

A network (also called a \emph{graph} in mathematics) comprises a set of nodes connected to one another by a set of edges. Street networks can be conceptualized as primal, directed, nonplanar graphs. A primal street network represents intersections as nodes and street segments as edges. A directed network has directed edges: that is, edge $uv$ points one-way from node $u$ to node $v$, but there need not exist a reciprocal edge $vu$. A planar network can be represented in two dimensions with its edges intersecting only at nodes. Most street networks are nonplanar---due to grade-separated expressways, overpasses, bridges, and tunnels---but most quantitative studies of urban street networks represent them as planar (e.g., Buhl et al. 2006; Cardillo et al. 2006; Barthélemy and Flammini 2008; Masucci et al. 2009; Strano et al. 2013) for tractability because bridges and tunnels are reasonably uncommon in some cities. Planarity may reasonably model the street networks of old European town centers, but poorly models the street network in more modern cities like Los Angeles with their grade-separated expressways, bridges, and tunnels. Thus, a planar simplification could yield inaccurate analyses.

This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales \cite{boeing_osmnx:_2017}. OSMnx is a Python-based research tool that easily downloads OpenStreetMap data for any place in the world and constructs it into a spatially-embedded graph-theoretic object for analysis and visualization. OpenStreetMap is a collaborative worldwide mapping project that makes its spatial data available via various APIs. These data are of high quality and compare well to CIA World Factbook estimates and US Census TIGER/Line data (Haklay 2010; Over et al. 2010; Zielstra and Hochmair 2011; Maron 2015). In 2007, OpenStreetMap imported the 2005 TIGER/Line roads data, and since then many community-led corrections and improvements have been made (Willis 2008). Many of these additions go beyond TIGER/Line's scope, including passageways between buildings, footpaths through parks, bike paths, and detailed feature attributes, such as finer-grained street classifiers, speed limits, etc.

To define the study sites and their spatial boundaries, this study uses three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 national TIGER/Line shapefile of US census bureau urban areas, released 19 August 2016. Census-defined urban areas comprise a set of census tracts that meet a minimum density threshold \cite{u.s._census_bureau_2010_2010}. In particular, we retain only the urbanized areas subset in this data set (i.e., areas with greater than 50,000 population), discarding the smaller urban clusters subset. The second set of geometries defines the municipal-scale study sites using 51 TIGER/Line shapefiles (again, 2016) of US census bureau places for all 50 states plus the District of Columbia. In particular, we discard the subset of census-designated places in this data set, instead retaining all the incorporated cities and towns in the United States.

The third and final set of geometries defines the neighborhood-scale study sites using 42 Zillow neighborhood shapefiles. These shapefiles contain the geometries of neighborhoods in 41 states plus the District of Columbia. This is a fairly new data set comprising nearly 7,000 neighborhoods in large American cities, but as \citet{schernthanner_spatial_2016} point out, Zillow does not publish the methodology it uses to construct these geometries. However, despite its newness it already has a track record in the academic literature. For instance, \citet{besbris_effect_2015} used Zillow boundaries to examine neighborhood stigma, and \citet{albrecht_indicator_2014} used these data to study neighborhood-level poverty in New York.

For all of these geometries, we use OSMnx to download the (drivable, public) street network contained within the geometry. First it buffers each geometry by 0.5 km, then downloads the street network within this buffered geometry. Next it constructs a street network from this data, corrects the topology, calculates street counts per node (this ensures that intersections are not considered cul-de-sacs simply because an incident edge connects to a node outside the desired polygon), then truncates the network to the original, desired polygon. OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it automatically calculates metric and topological measures for each. Such measures extend the toolkit commonly used by urban morphologists (Talen 2003). Brief descriptions of these measures are discussed here, but extended definitions and algorithms can be found in Newman (2010) and Barthélemy (2011).

Average street length, the mean edge length in the undirected representation of the graph, serves as a linear proxy for block size and indicates how fine-grained or coarse-grained the network is. Node density is the number of nodes divided by the area covered by the network. Intersection density is the node density of the set of nodes with more than one street emanating from them (thus excluding dead-ends). The edge density is the sum of all edge lengths divided by the area, and the physical street density is the sum of all edges in the undirected representation of the graph divided by the area. These density measures all provide further indication of how fine-grained the network is. Finally, the average circuity represents the average ratio between an edge length and the straight-line distance between the two nodes it links.

The average node degree, or mean number of edges incident to each node, quantifies how well the nodes are connected on average. Similarly, but more concretely, the average streets per node measures the mean number of physical streets (i.e., edges in the undirected representation of the graph) that emanate from each intersection and dead-end. This adapts the average node degree for physical form rather than directed circulation. The statistical and spatial distributions of number of streets per intersection characterize the type, prevalence, and dispersion of intersection connectedness in the network. Connectivity measures the minimum number of nodes or edges that must be removed from a connected graph to disconnect it. The average node connectivity of a network---the mean number of internally node-disjoint paths between each pair of nodes in the graph---represents the expected number of nodes that must be removed to disconnect a randomly selected pair of non-adjacent nodes \cite{beineke_average_2002}.

The clustering coefficient of a node is the ratio of the number of edges between its neighbors to the maximum possible number of edges that could exist between these neighbors. The weighted clustering coefficient weights this ratio by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the nodes in the network. Centrality indicates the importance of nodes in a network. Betweenness centrality evaluates the number of shortest paths that pass through each node \cite{barthelemy_betweenness_2004}. The maximum betweenness centrality in a network specifies the proportion of shortest paths that pass through the most important node. This is an indicator of resilience: networks with a high maximum betweenness centrality are more prone to failure or inefficiency should this single choke point fail. Closeness centrality ranks nodes as more central if they are on average closer to all other nodes. Finally, PageRank ranks nodes based on the structure of incoming links and the rank of the source node.

@@

\section{Analysis of Metropolitan-Scale Street Networks}

@@

\begin{table}
\centering
\caption{Central tendency and statistical dispersion for selected measures of all US urbanized areas' street networks.}
\label{table01}
\begin{tabular}{llllll}
	\toprule
	measure                                     & mean           & std dev        & min            & median         & max            \\
	\midrule
	Area (km2)                                  & 460.657        & 858.125        & 25.685         & 184.898        & 8937.429       \\
	Avg of the avg neighborhood degree          & 2.886          & 0.109          & 2.626          & 2.875          & 3.228          \\
	Avg of the avg weighted neighborhood degree & 0.032          & 0.018          & 0.021          & 0.03           & 0.321          \\
	Avg circuity                                & 1.076          & 0.019          & 1.023          & 1.074          & 1.14           \\
	Avg clustering coefficient                  & 0.042          & 0.009          & 0.015          & 0.042          & 0.071          \\
	Avg weighted clustering coefficient         & 0.002          & 0.001          & \textless0.001 & 0.001          & 0.006          \\
	Intersection count                          & 12582          & 26054          & 751            & 4593           & 307848         \\
	Avg degree centrality                       & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.007          \\
	Edge density (km/km2)                       & 13.455         & 2.137          & 7.961          & 13.352         & 21.233         \\
	Avg edge length (m)                         & 158.588        & 17.653         & 117.341        & 157.332        & 223.08         \\
	Total edge length (km)                      & 6353           & 12625          & 427            & 2393           & 1.42E+08       \\
	Proportion of dead-ends                     & 0.213          & 0.055          & 0.077          & 0.207          & 0.416          \\
	Proportion of 3-way intersections           & 0.593          & 0.046          & 0.444          & 0.591          & 0.778          \\
	Proportion of 4-way intersections           & 0.187          & 0.063          & 0.054          & 0.178          & 0.422          \\
	Intersection density (per km2)              & 26.469         & 6.256          & 12.469         & 26.029         & 49.423         \\
	Average node degree                         & 5.153          & 0.302          & 4.307          & 5.143          & 6.056          \\
	m                                           & 40890          & 83678          & 2516           & 14955          & 981646         \\
	n                                           & 16032          & 32585          & 874            & 5830           & 373309         \\
	Node density (per km2)                      & 33.628         & 7.641          & 17.675         & 33.071         & 61.655         \\
	Max PageRank value                          & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.003          \\
	Min PageRank value                          & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 \\
	Self-loop proportion                        & 0.008          & 0.008          & \textless0.001 & 0.006          & 0.071          \\
	Street density (km/km2)                     & 7.262          & 1.221          & 4.217          & 7.171          & 11.797         \\
	Average street segment length (m)           & 161.331        & 17.765         & 119.573        & 160.288        & 225.92         \\
	Total street length (km)                    & 3480           & 7026           & 222            & 1269           & 79046          \\
	Street segment count                        & 22011          & 45725          & 1281           & 7868           & 533757         \\
	Average streets per node                    & 2.764          & 0.162          & 2.223          & 2.77           & 3.217          \\
	\bottomrule
\end{tabular}
\end{table}

@@

urban area core city & land area km2 & avg circuity & avg clustering coefficient & dead-end ratio & 3-way ratio & 4-way ratio & intersect density/ km2 & street density km/km2 & avg street length (m) & avg streets/ node \\

@@

These metropolitan-scale analyses consider trends in the built form at the scale of broad self-organized human systems and urbanized regions. However, they aggregate multiple heterogenous neighborhoods and municipalities---the scales of human life, urban design projects, and planning jurisdiction---into single units of analysis. To disaggregate and analyze finer characteristics, the following sections examine municipal- and neighborhood-scale street networks across the United States.

\section{Analysis of Municipal-Scale Street Networks}

Table \ref{table03} shows similarly great variation in street network characteristics across the entire data set of 19,655 cities and towns. Again, this data set comprises the street networks of every incorporated city and town in the United States. Following recent work by \citet{barthelemy_modeling_2008} and \citet{strano_urban_2013}, we examine the relationship between the total street length $L$ and the number of nodes $n$. The former proposed a model of cities in which $L$ and $n$ scale nonlinearly as $n^{1/2}$, and the latter recently confirmed this empirically with a small sample of ten European cities' street networks. However, their small sample size may limit the generalizability and interpretability of their finding. To investigate this empirically, we examine the relationship between the total street length $L$ and the number of nodes $n$ for 19,655 US cities and towns and find a strong \emph{linear} relationship ($r^{2}=0.98$), as depicted in Figure \ref{fig03}, contradicting Strano et al. We also find a similar linear relationship at the metropolitan and neighborhood scales.

>>>>

\usepackage{graphicx}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{pdflscape}
\usepackage[USenglish]{babel}
\usepackage{csquotes}
\usepackage[numbers]{natbib}
\usepackage[affil-it]{authblk}

\bibliographystyle{chicago}

@@

OpenStreetMap offers an underexplored source of worldwide geospatial data useful to urban researchers. This paper uses the OSMnx software to automatically download and analyze over 27,000 street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---namely, every urbanized area, incorporated city and town, and Zillow neighborhood in the US. In turn it presents wide-ranging empirical findings on US urban form and street network characteristics. This study conducts its analysis of with extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. This analysis emphasizes urban street network measures relevant to urban design and morphology including structure, connectedness, centrality, and robustness. In total it cross-sectionally analyzes the street networks of 497 urbanized areas, 19,655 cities and towns, and 6,857 neighborhoods. These sample sizes are larger than those in any previous study and illustrate the use of OSMnx and OpenStreetMap as a new urban form research platform.

@@

On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States' Midwest and Great Plains available for free to all applicants \cite{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 270 million acres of public land (10\% of the entire US landmass) to private owners in the form of 1.6 million homesteads \cite{lee_kansas_1979, sherraden_inclusion_2005}. Towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \cite{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks.

A cross-sectional analysis of American urban form can reveal these artifacts and histories through street networks at metropolitan, municipal, and neighborhood scales. Network analysis is a natural approach to the study of cities as complex systems \cite{masucci_random_2009}. The empirical literature on street network analysis is growing ever richer, but suffers from some limitations---discussed in detail in \cite{boeing_osmnx:_2017} and summarized here. First, sample sizes have tended to be fairly small due to data availability, gathering, and processing constraints: most studies in the street network literature that conduct topological and/or metric analysis tend to have sample sizes ranging around 10 to 50 networks \cite{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, strano_urban_2013}, which may limit the generalizability and interpretability of findings. Second, reproducibility has been difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported \citep[e.g.,][]{porta_network_2006, strano_urban_2013}. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability \cite[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009}, or use dual graphs despite the loss of geographic and metric information \cite{batty_network_2005, jiang_integration_2002, ratti_space_2004, crucitti_centrality_2006, crucitti_centrality_2006-1}.

This study addresses these limitations by conducting an urban morphological analysis of street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. In particular, it examines urban street networks---represented here as primal, nonplanar, weighted multidigraphs with possible self-loops---focusing on structure, connectedness, centrality, and resilience. This study analyzes over 27,000 urban street networks at multiple overlapping scales across the US. Namely, it examines the street networks of every incorporated city and town (population size $N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$) in the US. To do so, it uses OSMnx---a new street network research toolkit \cite{boeing_osmnx:_2017}---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales.

The purpose of this paper is threefold. First, it describes and demonstrates a new methodology for easily and consistently acquiring, constructing, and analyzing large samples of street networks as nonplanar directed graphs. Second, it presents empirical findings of descriptive urban morphology for the street networks of every city, urbanized area, and Zillow neighborhood in the US. Third, it investigates with large sample sizes some claims previously made in the research literature with small sample sizes. This paper is organized as follows. In the next section, it discusses the data sources, tools, and methods used to collect, construct, and analyze these street networks. Next, it presents findings of the analyses at metropolitan, municipal, and neighborhood spatial scales. Finally, it concludes with a discussion of these findings and their implications for street network analysis, urban morphology, and city planning.

\section{Data and Methods}

A network (also called a \emph{graph} in mathematics) comprises a set of nodes connected to one another by a set of edges. Street networks can be conceptualized as primal, directed, nonplanar graphs. A primal street network represents intersections as nodes and street segments as edges. A directed network has directed edges: that is, edge $uv$ points one-way from node $u$ to node $v$, but there need not exist a reciprocal edge $vu$. A planar network can be represented in two dimensions with its edges intersecting only at nodes \cite{viana_simplicity_2013, fischer_spatial_2014}. Most street networks are nonplanar---due to grade-separated expressways, overpasses, bridges, tunnels, etc.---but most quantitative studies of urban street networks represent them as planar \cite{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009, strano_urban_2013} for tractability because bridges and tunnels are reasonably uncommon in some cities. Planar graphs may reasonably model the street networks of old European town centers, but poorly model the street networks in modern autocentric cities like Los Angeles or Tokyo with many grade-separated expressways, bridges, and underpasses.

\subsection{Study Sites and Data Acquisition}

This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales. OSMnx is a Python-based research tool that easily downloads OpenStreetMap data for any place in the world and constructs it into a spatially-embedded graph-theoretic object for analysis and visualization \cite{boeing_osmnx:_2017}. OpenStreetMap is a collaborative worldwide mapping project that makes its spatial data available via various APIs \cite{corcoran_analysing_2013, jokar_arsanjani_openstreetmap_2015}. Its data are of high quality and compare favorably to CIA World Factbook estimates and US Census TIGER/Line data \cite{haklay_how_2010, over_generating_2010, zielstra_comparative_2011, maron_how_2015, wu_improving_2005, frizzelle_importance_2009}. In 2007, OpenStreetMap imported the TIGER/Line roads (2005 vintage), and since then many community-led corrections and improvements have been made \cite{willis_openstreetmap_2008}. Many of these additions go beyond TIGER/Line's scope, including passageways between buildings, footpaths through parks, bike routes, and detailed feature attributes such as finer-grained street classifiers, speed limits, etc.

To define the study sites and their spatial boundaries, this study uses three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 national TIGER/Line shapefile of US census bureau urban areas. Each census-defined urban areas comprises a set of tracts that meet a minimum density threshold \cite{u.s._census_bureau_2010_2010}. We retain only the \emph{urbanized areas} subset of these data (i.e., areas with greater than 50,000 population), discarding the small \emph{urban clusters} subset. The second set of geometries defines our municipal-scale study sites using 51 separate TIGER/Line shapefiles (again, 2016) of US census bureau \emph{places} within all 50 states plus the District of Columbia. We discard the subset of \emph{census-designated places} (i.e., small unincorporated communities) in these data, while retaining every incorporated city and town in the US. The third set of geometries defines the neighborhood-scale study sites using 42 separate Zillow neighborhood shapefiles. These shapefiles contain neighborhood boundaries for major cities in 41 states plus the District of Columbia. This is a fairly new data set comprising nearly 7,000 neighborhoods in large American cities, but as \citet{schernthanner_spatial_2016} point out, Zillow does not publish the methodology it uses to construct these boundaries. However, despite its newness it already has a track record in the academic literature: \citet{besbris_effect_2015} used Zillow boundaries to examine neighborhood stigma and \citet{albrecht_indicator_2014} used them to study neighborhood-level poverty in New York.

For all of these geometries, we use OSMnx to download the (drivable, public) street network contained within the geometry, a process described in detail in \cite{boeing_osmnx:_2017} and summarized here. First OSMnx buffers each geometry by 0.5 km, then downloads the OpenStreetMap \enquote{nodes} and \enquote{ways} within this buffered geometry. Next it constructs a street network from these data, corrects the topology, calculates street counts per node, then truncates the network to the original, desired polygon (ensuring that intersections are not considered cul-de-sacs simply because an incident edge connects to a node outside the desired polygon). OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it calculates metric and topological measures for each network. Such measures extend the toolkit commonly used in urban form studies \cite{talen_measuring_2003, ewing_travel_2010}.

\subsection{Street Network Measures}

Brief descriptions of these measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. Average street length, the mean edge length in the undirected representation of the graph, serves as a linear proxy for block size and indicates how fine-grained or coarse-grained the network is. Node density is the number of nodes divided by the area covered by the network. Intersection density is the node density of the set of nodes with more than one street emanating from them (thus excluding dead-ends). The edge density is the sum of all edge lengths divided by the area, and the physical street density is the sum of all edges in the undirected representation of the graph divided by the area. These density measures all provide further indication of how fine-grained the network is. Finally, the average circuity represents the average ratio between an edge length and the straight-line distance between the two nodes it links.

The average node degree, or mean number of edges incident to each node, quantifies how well the nodes are connected on average. Similarly, but more concretely, the average streets per node measures the mean number of physical streets (i.e., edges in the undirected representation of the graph) that emanate from each intersection and dead-end. This adapts the average node degree for physical form rather than directed circulation. The statistical and spatial distributions of number of streets per intersection characterize the type, prevalence, and dispersion of intersection connectedness in the network. Connectivity measures the minimum number of nodes or edges that must be removed from a connected graph to disconnect it. The average node connectivity of a network---the mean number of internally node-disjoint paths between each pair of nodes in the graph---represents the expected number of nodes that must be removed to disconnect a randomly selected pair of non-adjacent nodes \cite{beineke_average_2002,dankelmann_bounds_2003}.

The clustering coefficient of a node is the ratio of the number of edges between its neighbors to the maximum possible number of edges that could exist between these neighbors \cite{jiang_topological_2004, opsahl_clustering_2009}. The weighted clustering coefficient weights this ratio by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the nodes in the network. Centrality indicates the importance of nodes in a network \cite{huang_trajgraph:_2016, zhong_revealing_2017}. Betweenness centrality evaluates the number of shortest paths that pass through each node \cite{barthelemy_betweenness_2004}. The maximum betweenness centrality in a network specifies the proportion of shortest paths that pass through the most important node. This is an indicator of resilience: networks with a high maximum betweenness centrality are more prone to failure or inefficiency should this single choke point fail. Closeness centrality ranks nodes as more central if they are on average closer to all other nodes. Finally, PageRank ranks nodes based on the structure of incoming links and the rank of the source node \cite{brin_anatomy_1998, jiang_predicting_2008, agryzkov_algorithm_2012, chin_geographically_2015, gleich_pagerank_2015}.

@@

\section{Results}

\subsection{Analysis of Metropolitan-Scale Street Networks}

@@

\caption{Central tendency and statistical dispersion for selected measures of all US urbanized areas' street networks.}
\label{table01}
\begin{tabular}{llllll}
	measure                                     & mean           & std dev        & min            & median         & max            \\
	Area (km\textsuperscript{2})                                  & 460.657        & 858.125        & 25.685         & 184.898        & 8937.429       \\
	Avg of the avg neighborhood degree          & 2.886          & 0.109          & 2.626          & 2.875          & 3.228          \\
	Avg of the avg weighted neighborhood degree & 0.032          & 0.018          & 0.021          & 0.03           & 0.321          \\
	Avg circuity                                & 1.076          & 0.019          & 1.023          & 1.074          & 1.14           \\
	Avg clustering coefficient                  & 0.042          & 0.009          & 0.015          & 0.042          & 0.071          \\
	Avg weighted clustering coefficient         & 0.002          & 0.001          & \textless0.001 & 0.001          & 0.006          \\
	Intersection count                          & 12582          & 26054          & 751            & 4593           & 307848         \\
	Avg degree centrality                       & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.007          \\
	Edge density (km/km\textsuperscript{2})                       & 13.455         & 2.137          & 7.961          & 13.352         & 21.233         \\
	Avg edge length (m)                         & 158.588        & 17.653         & 117.341        & 157.332        & 223.08         \\
	Total edge length (km)                      & 6353           & 12625          & 427            & 2393           & 1.42E+08       \\
	Proportion of dead-ends                     & 0.213          & 0.055          & 0.077          & 0.207          & 0.416          \\
	Proportion of 3-way intersections           & 0.593          & 0.046          & 0.444          & 0.591          & 0.778          \\
	Proportion of 4-way intersections           & 0.187          & 0.063          & 0.054          & 0.178          & 0.422          \\
	Intersection density (per km\textsuperscript{2})              & 26.469         & 6.256          & 12.469         & 26.029         & 49.423         \\
	Average node degree                         & 5.153          & 0.302          & 4.307          & 5.143          & 6.056          \\
	m                                           & 40890          & 83678          & 2516           & 14955          & 981646         \\
	n                                           & 16032          & 32585          & 874            & 5830           & 373309         \\
	Node density (per km\textsuperscript{2})                      & 33.628         & 7.641          & 17.675         & 33.071         & 61.655         \\
	Max PageRank value                          & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.003          \\
	Min PageRank value                          & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 \\
	Self-loop proportion                        & 0.008          & 0.008          & \textless0.001 & 0.006          & 0.071          \\
	Street density (km/km\textsuperscript{2})                     & 7.262          & 1.221          & 4.217          & 7.171          & 11.797         \\
	Average street segment length (m)           & 161.331        & 17.765         & 119.573        & 160.288        & 225.92         \\
	Total street length (km)                    & 3480           & 7026           & 222            & 1269           & 79046          \\
	Street segment count                        & 22011          & 45725          & 1281           & 7868           & 533757         \\
	Average streets per node                    & 2.764          & 0.162          & 2.223          & 2.77           & 3.217          \\

@@

urban area core city & land area km\textsuperscript{2} & avg circuity & avg clustering coefficient & dead-end ratio & 3-way ratio & 4-way ratio & intersect density/ km\textsuperscript{2} & street density km/km\textsuperscript{2} & avg street length (m) & avg streets/ node \\

@@

These metropolitan-scale analyses consider trends in the built form at the scale of broad self-organized human systems and urbanized regions. However, they aggregate multiple heterogenous neighborhoods and municipalities---the scales of human life, urban design projects, and planning jurisdiction---into single units of analysis. To disaggregate and analyze finer characteristics, the following sections examine municipal- and neighborhood-scale street networks across the US.

\subsection{Analysis of Municipal-Scale Street Networks}

Table \ref{table03} shows similarly great variation in street network characteristics across the entire data set of 19,655 cities and towns. Again, this data set comprises the street networks of every incorporated city and town in the US. Following recent work by \citet{barthelemy_modeling_2008} and \citet{strano_urban_2013}, we examine the relationship between the total street length $L$ and the number of nodes $n$. The former proposed a model of cities in which $L$ and $n$ scale nonlinearly as $n^{1/2}$, and the latter recently confirmed this empirically with a small sample of ten European cities' street networks. However, their small sample size may limit the generalizability and interpretability of their finding. To investigate this empirically, we examine the relationship between the total street length $L$ and the number of nodes $n$ for 19,655 US cities and towns and find a strong \emph{linear} relationship ($r^{2}=0.98$), as depicted in Figure \ref{fig03}, contradicting Strano et al. We also find a similar linear relationship at the metropolitan and neighborhood scales.

\begin{table}
\centering
\caption{Selected summary stats for every incorporated city and town in the US.}
\label{table03}
\begin{tabular}{llllll}
	\toprule
	measure                                     & mean    & std dev & min            & median         & max       \\
	\midrule
	Area (km\textsuperscript{2})                                  & 16.703  & 107.499 & 0.039          & 3.918          & 7434.258  \\
	Avg of the avg neighborhood degree          & 2.94    & 0.297   & 0.4            & 2.953          & 3.735     \\
	Avg of the avg weighted neighborhood degree & 0.033   & 0.141   & \textless0.001 & 0.029          & 9.357     \\
	Avg circuity                                & 1.067   & 0.159   & 1              & 1.055          & 20.452    \\
	Avg clustering coefficient                  & 0.048   & 0.041   & \textless0.001 & 0.04           & 1         \\
	Avg weighted clustering coefficient         & 0.01    & 0.018   & \textless0.001 & 0.005          & 0.524     \\
	Intersection count                          & 324     & 1266    & 0              & 83             & 62996     \\
	Avg degree centrality                       & 0.093   & 0.136   & \textless0.001 & 0.052          & 2.667     \\
	Edge density (km/km\textsuperscript{2})                       & 12.654  & 6.705   & 0.006          & 11.814         & 58.603    \\
	Avg edge length (m)                         & 161.184 & 80.769  & 25.822         & 144.447        & 3036.957  \\
	Total edge length (km)                      & 159.067 & 578.521 & 0.052          & 40.986         & 24728.326 \\
	Proportion of dead-ends                     & 0.192   & 0.093   & \textless0.001 & 0.184          & 1         \\
	Proportion of 3-way intersections           & 0.572   & 0.11    & \textless0.001 & 0.579          & 1         \\
	Proportion of 4-way intersections           & 0.237   & 0.129   & \textless0.001 & 0.217          & 1         \\
	Intersection density (per km\textsuperscript{2})              & 29.363  & 21.607  & \textless0.001 & 24.719         & 259.647   \\
	Average node degree                         & 5.251   & 0.668   & 0.8            & 5.268          & 7.166     \\
	m                                           & 1046    & 3924    & 2              & 275            & 176161    \\
	n                                           & 401     & 1516    & 2              & 103            & 71993     \\
	Node density (per km\textsuperscript{2})                      & 35.449  & 24.409  & 0.047          & 30.718         & 296.74    \\
	Max PageRank value                          & 0.034   & 0.046   & \textless0.001 & 0.021          & 0.87      \\
	Min PageRank value                          & 0.005   & 0.018   & \textless0.001 & 0.002          & 0.5       \\
	Self-loop proportion                        & 0.005   & 0.015   & \textless0.001 & \textless0.001 & 1         \\
	Street density (km/km\textsuperscript{2})                     & 6.528   & 3.435   & 0.003          & 6.109          & 29.302    \\
	Average street segment length (m)           & 162.408 & 81.035  & 25.822         & 145.479        & 3036.957  \\
	Total street length (km)                    & 86.096  & 331.048 & 0.026          & 21.005         & 15348.008 \\
	Street segment count                        & 558     & 2208    & 1              & 140            & 107393    \\
	Average streets per node                    & 2.851   & 0.282   & 1              & 2.852          & 4         \\
	\bottomrule
\end{tabular}
\end{table}
